\providecommand{\main}{.}
\documentclass[12pt,a4paper]{report} %Modalità draft settata qui

\input{\main/header.tex}

\begin{document}

\newgeometry{total={170mm,257mm}, left=20mm, top=20mm}
%Maybe add some space there
\begin{center}
                \line (1,0){350} \\
                \huge{\bfseries Statistical Mechanics of Complex Systems}\\
                [2mm]
                \textsc{\normalsize Prof. \textit{Amos Maritan}}
                \vspace{-0.5em}\\
                \textsc{\normalsize University of Padova}\\
                \vspace{-1em}
                \line (1,0){350} \\
        [0.2cm]
        \textsc{\normalsize Written by: \textit{Francesco Manzali}}\\
                \textsc{\normalsize Academic Year 2019/20}\\ 
        {\scriptsize Compiled on \today}\\
        [0.2cm]
        {\scriptsize \doclicenseLongText}
\end{center}

%---Frontpage End---%

\tableofcontents 
\addtocontents{toc}{\protect\thispagestyle{empty}}
\pagenumbering{gobble}

\clearpage
\section*{Introduction}
In the following, you will find the notes of \q{Statistical Mechanics of Complex Systems}. As of A.A. 2019-20, the course is divided in two parts:
\begin{itemize}
        \item \textbf{Part A}, dealing with mathematical methods, the basics of statistical mechanics and stochastic processes (Brownian motion).
        \item \textbf{Part B}, focusing on \textit{entropy}, the Ising Model, variational methods and dynamics of complex systems.
\end{itemize}
Notes of part B are \textit{complete}, in the sense that all lectures and topics are covered. However, pay attention as they may well be mistakes, typos and other not-so-good stuff.

On the other hand, part A, which was reserved to students of mathematical engineering, contains only the transcriptions of the first few lectures. All the missing required topics can be found, in much more detail, in the first chapters of the following notes from Physics of Data courses:
\begin{itemize}
        \item \q{Statistical Mechanics} by Alice Pagano: \href{https://github.com/AlicePagano/Lectures-Statistical-Mechanics}{link}.
        \item \q{Models of Theoretical Physics} (Brownian motion) by Francesco Manzali: \href{https://www.overleaf.com/read/hwzmdmrmmhzn}{link}.
\end{itemize}

Many intermediate steps and additional information have been added to make the notes more clear, at the cost of a bit of redundancy.

If you find any mistake/typo/missing thing, please write at \textit{francesco.manzali@studenti.unipd.it} (or even just for feedback).

\begin{flushright}
\textit{Francesco Manzali}, 02/07/2019
\end{flushright}

\begin{flushright}
\textit{The contract is complete. Your wish has surpassed entropy. Now release your new power!}\\
-Kyubey - Puella Magi Madoka Magica
\end{flushright}

\subsection*{Aknowledgments}
Thanks to Alice Pagano for providing corrections and realizing several figures in tikz. Thanks to Karan K.H.M., Marco Ballarin (and everyone that I am regretfully forgetting) for all the feedback on typos and exercises.

\clearpage
            
\restoregeometry

\chapter{Statistical Mechanics}

%---First part (3CFU)---%
\subfile{Lectures/Maritan/09_03_20.tex}
\subfile{Lectures/Maritan/11_03_20.tex}
\subfile{Lectures/Maritan/12_03_20.tex}
\subfile{Lectures/Maritan/16_03_20.tex}

\danger 
\begin{expl}
        \textbf{Missing lectures}. The following $9$ lectures are missing, ranging from 18/03 to 06/04.  Their topics can be found at the following \href{https://docs.google.com/spreadsheets/d/1H_VqhfWnARh3Nlv6bU2MoWXACttQx0YhCds1guY-hKE/edit?usp=sharing}{link}.
\end{expl}

%TO Review:
%\subfile{Lectures/Maritan/18_03_20.tex}
%\subfile{Lectures/Maritan/19_03_20.tex}

%TO COPY:
%\subfile{Lectures/Maritan/23_03_20.tex}
%\subfile{Lectures/Maritan/25_03_20.tex}
%\subfile{Lectures/Maritan/26_03_20.tex}
%\subfile{Lectures/Maritan/30_03_20.tex}
%\subfile{Lectures/Maritan/01_04_20.tex}
%\subfile{Lectures/Maritan/02_04_20.tex}
%\subfile{Lectures/Maritan/06_04_20.tex}

%---Second part (6CFU)---%
\subfile{Lectures/Maritan/08_04_20.tex}
\subfile{Lectures/Maritan/09_04_20.tex}

%Finish review:
\subfile{Lectures/Maritan/15_04_20.tex}
\subfile{Lectures/Maritan/16_04_20.tex}
\subfile{Lectures/Maritan/20_04_20.tex}

\subfile{Lectures/Maritan/22_04_20.tex}
\subfile{Lectures/Maritan/23_04_20.tex}

%To copy:
\subfile{Lectures/Maritan/27_04_20.tex}
\subfile{Lectures/Maritan/29_04_20.tex}

%Suweiss part (to be copied):
%\subfile{Lectures/Suweiss/30_04_20.tex}
%\subfile{Lectures/Suweiss/04_05_20.tex}
%\subfile{Lectures/Suweiss/06_05_20.tex}
%\subfile{Lectures/Suweiss/07_05_20.tex}

%(Almost) finished
\subfile{Lectures/Maritan/11_05_20.tex}
\subfile{Lectures/Maritan/13_05_20.tex}
\subfile{Lectures/Maritan/14_05_20.tex}
\subfile{Lectures/Maritan/18_05_20.tex}
\subfile{Lectures/Maritan/20_05_20.tex}
\subfile{Lectures/Maritan/21_05_20.tex}
\subfile{Lectures/Maritan/25_05_20.tex}

\subfile{Lectures/Maritan/27_05_20.tex}
\subfile{Lectures/Maritan/28_05_20.tex}
\subfile{Lectures/Maritan/01_06_20.tex}




%---Bibliography---%
\clearpage
\printbibliography

%---Analytical Index---%
\printindex

\end{document}

